\documentclass[a4paper,twoside,titlepage]{article}
\usepackage[top=3cm,bottom=3cm,left=3cm,right=3cm]{geometry}
\usepackage{url}

\renewcommand\familydefault{\sfdefault}

\title{Advanced Filesystem Layout, AFSL Version 1}
\author{Magnus Achim Deininger}


\begin{document}
\maketitle
\tableofcontents

\newpage

\section*{Preface}
\addcontentsline{toc}{section}{Preface}
Current filesystem layouts employed by contemporary operating systems suffer
from a number of shortcomings. Among the more common of these are general
inconsistencies and issues with contemporary multi-architecture systems. Minor,
similar issues occur when trying to develop for architectures other than that of
the host system. In general, it is becoming increasingly common that binaries
(executables and libraries) of distinct systems occur together on a single
system. Examples for these would include contemporary desktop systems with
``64-bit'' capabilities mixing binaries compiled for x86-64 and x86-32 on the
same filesystem with poor separation, or average Linux installations containing
binaries intended for Microsoft Windows for use with WINE, or FreeBSD
installations including binaries intended for Linux.

This has lead to what may be perceived as inconsistencies in order to fix some
of these issues, such as the lib/lib32/lib64 dilemma that would require sane
software build scripts to figure out in which of these directories to put
created libraries. This alone is already inconsistent with requirements that
expressly forbid this -32/-64 extension on bin/ directories\footnote{See the FHS
document for details.}, and the idea itself seems rather hackish.

Additionally, the original reasoning behind a split of the root filesystem and
the /usr hierarchy (and also somewhat the /opt and /usr/local hierarchies) seems
to have been to be able to separate machine-local binaries and configuration
from site-local binaries and shared data and read-only from read-write data.
Originally a physical separation between the hierarchies was necessary, but due
to the ability of contemporary operating systems to unify multiple filesystem
hierarchies, sometimes called "union mounts", this division seems quite obsolete
\footnote{Operating Systems such as Plan 9 seem to agree with this.}.

This paper will try to explore an alternative filesystem hierarchy scheme that
should provide consistency for years to come. Keep in mind that this is a draft,
not yet a final document.

\newpage

\section{The Advanced Filesystem Layout}
The general idea behind this filesystem layout is fairly simple: "traditional"
directories are at the end of an arbitrarily long list of host specifications.
This leads to a general layout that looks something like this:
$os/architecture/subarchitecture/(bin|lib)$. The idea is to get more and more
specific along the path, with the operating system being considered less
specific than the hardware architecture. This latter choice is fairly random,
but it seems to be easier or more common to run binaries from a different
architecture on the same operating system than it is to run binaries from a
different operating system even when on the same architecture (although there
are legimitate examples of both).

\subsection{General Notes}
All of the elements that vary by target system are written $like this$. These
are always assumed to be lowercase, without spaces or underscores. Use dashes
(`-') as a delimiter when needed. For example, for $architecture$ use `x86-64'
instead of `x86\_64' or `power-macintosh' instead of `power macintosh'. Also,
use values that are consistent and generic across all architectures and
operating systems. That means, instead of `power-macintosh', what you really
should be using is `powerpc-32' or `powerpc-64', and instead of `amd64' or
`em64t' you should be using `x86-64'.

\subsection{Hierarchical Elements}
Paths discussed in this section are prefixes for other parts. Also refer to the
appendix for values to be used for common variables in this section.

\subsubsection{/($os$/)?($architecture$/)?($subarchitecture$/)*/: Architecture
Classification}
$os$ refers to a short code, hopefully a single word that would describe the
intended target operating systems. Examples for this would be `linux' or
`plan9'. $architecture$ is the general architecture class of the target host,
for example `x86-32', `x86-64', `powerpc-32', etc. Unless there is only one
conceivable implementation to refer to, this should always include the GPR or
address space length of the architecture. For both $os$ and $architecture$, the
value `generic' may be used if it doesn't matter which operating system or cpu
architecture some of the content is, but it would still be desirable to keep the
parent subdirectory count low and this could be achieved by means of a
placeholder such as `generic'.

$subarchitecture$ is anything that could be used to further specify requirements
that the binaries may have. This may be anything as specific as `sse3' or
`altivec', or something more generic such as `i686' or just `little-endian'.
System vendor specifications may also be used here, i.e. `sun', `intel',
`apple', etc.

As should be evident by the title above, more than one $subarchitecture$
specification may be used. They must still be specified in reverse order of
importance, that is, more generic specifications must come first, whereas
more specific details should be last. This is to faciliate finding applicable
files in the hierarchy.

With all of these, redundant specifications may be omitted. For example, if the
target system will always be a linux system, then it's allowed to drop the
/linux prefix.

Now, when actually locating applicable files or hierarchies, it should also be
noted that more specific, applicable subhierarchies always take precedence over
less specific but nevertheless applicable hierarchies. That is to say, files in
`/linux/x86-32/sse2/bin' take precedence over `/linux/x86-32/bin' or even just
`/bin'. This makes optimisations a lot more effective and easy, by being able to
put binaries that are more heavily optimised for a certain cpu model in the
filesystem without interfering with the other binaries on the system.

\subsubsection{/medium/$mediumname$[0-9]*/: Storage Media}
Mountpoints for storage media belong under this

\subsubsection{/home/$username$/: Home Directories}
You may use the architecture classification as discussed above in subdirectories
below here if necessary.

\subsubsection{/service/$servicename$/: Services}
You may use the architecture classification as discussed above in subdirectories
below here if necessary.

\subsubsection{/dev, /proc, /sys, /net, (...): Special Kernel Filesystems}
These obviously have operating-system specific meaning and layout. They usually
need to be kept in place (for example sysfs on Linux is specifically documented
as being required to be mounted on /sys). Obviously only nodes that are actually
going to be mounted need to be created on a filesystem, but having more than
necessary doesn't seem to hurt, so it's unspecified whether or not unavailable
special filesystems still have their respective directories created or not.

\subsection{Final Path Elements}
Paths in this section are the ending parts of a sequence of path elements.

\subsubsection{.+/bin: Binaries}
All binaries (`executables') go into this directory. No exceptions, not even for
``system binaries'', which would traditionally end up in /sbin. If you need
finer-grained control or systematic clustering of related binaries, use a
subdirectory in /service instead.

\subsubsection{.+/lib: Libraries}

\subsubsection{.+/configuration or .+/etc: Configuration Files}

\subsubsection{.+/documentation or .+/doc: Documentation}

\subsubsection{.+/temporary or .+/tmp: Temporary Storage}

\subsubsection{.+/work: Intermediary Storage}

\subsection{Notable Differences to the FHS}

\subsubsection{.+/sbin vs .+/bin}
/sbin is pointless: system binaries may sound like a nice thing to hide from
users, but deciding which of the binaries are system binaries and which are
regular binaries is fairly arbitrary. For example the `ifconfig' programme is
often put in /sbin instead of /bin, although there are legitimate uses of
ifconfig for regular users. The programme's query functions immediately come to
mind. Since this functionality works fine for regular users, they can just
access the programme by entering the complete path, so the ``hiding'' was both
point- and fruitless.

Additionally the split is nearly impossible to perform systematically on
operating systems that have finer-grained access control than a simple root/user
divide. Operating systems like Plan9 or Linux\footnote{This may not seem obvious
at first, but internally the Linux kernel is documented to use so-called
`capabilities' that provide fine-grained access control to privileged functions}
come to mind. With these operating systems, half the users would probably end up
with /sbin in their path anyway since they'd have legitimate use for some system
programme or another.

For these reasons, there is no /sbin with this layout. If you really want to
hide system binaries, use filesystem namespaces and union mounts instead, but
remember that hiding them really is pointless.

\subsubsection{/var/tmp vs /work}

\subsubsection{/media vs /medium}
/media was quite an odd inconsistency. Virtually everything else used words or
abbreviations in the singular form, this is about the only place where a plural
was used. The idea to split it up was good though, so there we go.

\subsubsection{/var, /opt and /usr}

\newpage

\section{Usage}

\subsection{Environment Variables}

\newpage

\section{Issues with Legacy Systems}
There are some obvious incompatibilities with legacy filesystem layouts, at
least when employing the more descriptive paths.

\subsection{POSIX Compatibility Considerations}
The most notable problem to be encountered with live multi-host setups is the
likely lack of an appropriate /bin/sh binary. It should be possible to
circumvent this issue with either automatically created symlinks on boot or
with strategic application of namespaces.

It should be noted that this problem is easily be underestimated. Current BSD
and Linux systems, for example, employ a large amount of shell scripting for
common functionality, including things such as the ``init scripts'' that are
responsible for bringing up system services. In these script files, the first
line starts with a shabang, followed by the path to a script interpreter. Since
POSIX guarantees that /bin/sh is a valid shell script interpreter, scripts `out
in the wild' will rely on this and reference this location. The same applies to
the system() function in the standard C library.

\subsection{Booting}
During bootup the kernel needs to find and execute any binaries on the root
filesystem that are required to boot into userspace. With most other filesystem
layouts, this is easily done, as the binaries were usually always in the same,
so it's just a matter of hardcoding the location, however with this new layout
it's not quite that easy anymore, as the location of the binary might depend on
the host the computer is run on. (`might depend', because /bin/init would still
be a legitimate path with this filesystem layout.)

There would be a number of ways to go about this issue. One of them would be
by means of a small bootstrap partition or initramfs or similar with appropriate
binaries in a fixed location. This has the advantage of not requiring any
modifications to the kernel itself and it should always work. However, it'd be
quite a pain in the neck to maintain this separate, boot-only filesystem.

An alternative would be making the bootloader pass a parameter to the kernel
with the location of an appropriate init programme. This should be possible by
default with a good deal of popular kernels and bootloaders, such as FreeBSD or
Linux with GRUB or LILO, and is frequently done in practice with systems
employing ``alternative init systems''. Since the bootloader needs to know the
appropriate kernel to execute on the host system anyway, it is very likely that
it would be possible to specify a binary that would at least execute properly on
the target machine.

The kernel itself could also be modified to search for a suitable binary to
execute. This method is fairly invasive, as it requires direct modification of
kernel source code, but it would nevertheless be possible and provide a clean and
final solution to the issue.

\newpage

\begin{appendix}
\section{Path Elements}
The values listed here are to be considered mandatory to follow. If you don't
like some of them, you can still omit them most of the time, or suggest better
alternatives.

\subsection{$os$: Operating System Codes}

\subsection{$architecture$: CPU Architecture Codes}

\subsection{$subarchitecture$: CPU Sub-Architecture Codes}

\subsection{$mediumname$: Storage Medium Codes}

\subsection{$servicename$: Service Codes}

\end{appendix}

\newpage

\begin{thebibliography}{alpha}
\addcontentsline{toc}{section}{References}

\end{thebibliography}

\end{document}
